NATIONALIST BIAS IN SPORT PERFORMANCE EVALUATIONS

### 2019-08-12 revised-manuscript/response_to_reviewers.Rmd

> *3) Furthermore on page 12 it is stated that there would be a fluctuation in the number of observations due to missing values in the independent variables. Table 2 yet shows 18.860 observations for all three models presented, which indicates that all the available observations (3772 jumps x 5 judges) have been used to calculate the first three estimations.*

Thank you for pointing this out. This was an error on our part. It is corrected in the current version of the manuscript.

> *Home advantage is divided into a physical effect and a social effect. These constructs should be explained more clearly (see e.g. p. 10). I suspect that - according to the state of research (referenced on p. 5) - actual factors influencing performance exist (the sporting performance in the home country is actually on average higher than in a host country), but also influenced due to social pressure from the audience on the adjudicators, which can be assessed as higher in the home countries of the athletes than in the host countries (this is also explained on p. 10, this would then again be a distortion of judgement).*

In order to provide a more focused and coherent manuscript, we have removed the home field advantage aspect from this paper. Our decision was based on the fact that we expanded the nationalist bias analysis, and this expanded the scope of the work outside of a single article. We may proceed with a deeper analysis of home advantages in a subsequent paper.

If the Editor and the reviewers found much promise in the home field advantage component, we are of course willing to put it back in.

### 2019-08-12 revised-manuscript/manuscript.Rmd
Revising ongoing; now new analysis

---
title: "NATIONALIST AND HOME FIELD BIAS IN SPORT PERFORMANCE EVALUATIONS"
subtitle: "An example from the ski jumping World Cup"
runhead: Judging biases in skijumping
author:
- affiliation: Department of Medical Statistics, University of Oslo
  name: Leiv Tore Salte Rønneberg
- affiliation: Department of Sociology and Human Geography, University of Oslo.
  name: Torkild Hovde Lyngstad
- affiliation: European University Institute
  name: Juho Härkönen
date: "14 september 2017"
keywords: nationalist bias, home field bias, skijumping, judges
abstract: "High stakes and prizes in international sports competitions necessitate efficiently identifying legitimate winners. Illegitimate biases of non-independent judges and home field advantages represent two major threats to the legitimacy of international competitions. Judges’ nationalities represent one source of bias, as they may prefer athletes of their own nation. Judges may also favour athletes competing within their home countries, creating another potential bias. Such biases, conscious or unconscious, may affect the outcomes of competitions. The literature offers no complete consensus on the magnitude of such biases, hence not whether they affect competition outcomes or not. In this paper, we shed light on these problems, using international ski jumping as an example. We draw on data from the FIS World Cup competitions in the 2006-2008 and 2015-2016 seasons, and employ several fixed-effects based regression designs. Our results reaffirm suspicions of national bias and home field advantage in major ski jumping competitions, but also show their magnitude is too little to be of major relevance to competition outcomes. There is a trend towards smaller nationalist biases  "
thanks: "The authors are grateful for financial support from the Småforsk programme at the University of Oslo, and to Marte Knutsson for research assistance in the early phase of establishing the data set.  At publication time, all code and material used to produce this paper (with the exception of the proprietary data) will be available. The authors are also grateful to Matti Nykänen for being an enduring source of inspiration in many aspects of our lives."
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
    template: /home/torkildl/Dropbox/Setups/thl/rmd-xelatex-template-ms.tex
  html_document: default
bibliography: /home/torkildl/Dropbox/Literature/thl-main-library.bib
anonymous: yes
fontsize: 12pt
geometry: margin=2cm
header-includes:
    - \usepackage{placeins}
    - \usepackage{dcolumn}
    - \usepackage{tabularx}
    - \usepackage{lscape}
    - \usepackage{caption}
fig_caption: false
fig.align: "center"
fig.width: 5
---

@@

# Read in data
jumps <- as.data.frame(read_dta(here("data_prepared.dta"))) %>%
    select(order(names(.))) %>%
    arrange(jumpid)
tbl_df(jumps)

@@

# Correct judge names
pairs <- matrix(c("YERLY James","YERRLY James","ASPHAUG Kolbjoern","ASPHAUG Kolbjoer","DALLE AVE Sandro","DALLE Ave Sandro","EGLOFF Ernst","EGLOFF Emst","KARJALAINEN Pirjo","KARJALAINEN Pirvo","KUUSINEN Eero","KUUSINE Eero","LOENG Geir","LOENG Geir S.","LORENZ Michael","LORENTZ Michael","NYMAN Tom","NYMAN Torn","SCHMIEDER Jorg","SCHMIEDER Joerg","SCHMIEDERER Konrad","SCHIMIEDERER Konrad"), ncol=2, byrow=TRUE)
pairs[1,1]
for (x in 1:nrow(pairs)) {
    pair <- pairs[x,]
    j <- j %>% mutate(judge_name = ifelse(judge_name==pair[2],pair[1],judge_name))
}

# Various data preparation
d <- j %>% 
  # Correct erroneous nation coding
  mutate(judge_nation = ifelse(judge_nation=="JAP", "JPN", judge_nation)) %>%
  # Convert to factors
  mutate(judge_name = as.factor(judge_name)) %>%
  mutate(judge_nation = as.factor(judge_nation)) %>%
  mutate(country = as.factor(country)) %>%
  # Generate indicator of national bias
  mutate(natbias = ifelse(competitornation==judge_nation,1,0)) %>%
  # Generate indicator of own nation represented in judge panel 
  group_by(jumpid) %>%
  mutate(represented = max(natbias)) %>%
  ungroup %>%
  # and indicator of compansation bias by other judges
  mutate(compensating = represented - natbias) %>%
  # Generate home field advantage dummy
  mutate(athome = ifelse(competitornation==country,1,0)) %>%
  # Generate indicator of national biar per country
  mutate(countrybias = ifelse(natbias==1,as.character(judge_nation),"NON")) %>%
  mutate(countrybias = as.factor(countrybias)) %>%
  # Home field advantage variables
  mutate(athome = ifelse(competitornation==country,1,0)) %>%
  mutate(competitorname = as.factor(competitorname)) %>%
  mutate(competitionplace = as.factor(competitionplace)) %>%
  mutate(athseason = as.factor(str_c(competitorname,as.character(season)))) %>%
  # Drop unneeded variables
  select(everything())

#
# ESTIMATIONS: NATIONAL BIAS  MODELS
#
# Model 1, Naïve model of national bias
model1 <- lm(data = d,judge ~ natbias)
summary(model1)

# Model 2: introducing jump fixed effects
model2 <- felm(data = d, judge ~ natbias | jumpid)
summary(model2)

# Model 3: Introducing judge fixed effects
model3 <- felm(data = d, judge ~ natbias  | jumpid + judge_name)
summary(model3)

# Model 4: Does national bias vary among nations?
model4 <- felm(data = d, judge ~ relevel(countrybias,ref = "NON") | jumpid + judge_name)
summary(model4)

# 
# * HOME FIELD BIAS MODELS
# 
# Model 5: Length model
e <- filter(d, judgeno==1)
model5 <- felm(data=e, lenpts ~ athome + competitionplace | athseason)
summary(model5)

# Model 6: Style points model
model6 <- felm(data=d, judge ~ athome + competitionplace + judge_name + speed + length| athseason)
summary(model6)

#
# Season 2015-2016 analysis
#
season2016 <- data.table::fread(here("scrape/scraped-data-season-2016.csv"))

scores16 <- season2016 %>% 
    select(contains("judge"), competitorNation, country, place, speed, length, competitorName, gender) %>%
    mutate(jumpid = row_number()) %>%
    gather(key=jud, value=val, contains("judge")) %>%
    mutate(info = case_when(str_detect(jud, "Name") ~ "judge_name",
                            str_detect(jud,"Nation") ~ "judge_nation",
                            TRUE ~ "judge")) %>%
    mutate(ltr = str_sub(jud, start=-1)) %>%
    select(-jud) %>%
    group_by(jumpid, ltr) %>%
    arrange(jumpid, ltr) %>% 
    spread(key=info, value=val) %>%
    ungroup %>% 
    mutate(natbias = if_else(competitorNation==judge_nation,1,0)) %>%
    mutate_at(c("judge"), as.numeric) %>% 
    mutate(countrybias = factor(ifelse(natbias==1,as.character(judge_nation),"NON"))) %>% 
    mutate(competitionplace = str_sub(place, start=1,end=-6)) %>% 
    mutate(athome = if_else(country==competitorNation,1,0))

# NATIONAL BIAS MODELS WITH 2016 DATA
#
### Model 1b: Naïve model
model1b <- lm(data=scores16, judge ~ natbias)
summary(model1b)

### Model 2b: Introducing jump fixed effects
model2b <- felm(data = scores16, judge ~ natbias | jumpid)
summary(model2b)

### Model 3b: Introducing judge fixed effects
model3b <- felm(data = scores16, judge ~ natbias  | jumpid + judge_name)
summary(model3b)

# Model 4b: Does national bias vary among nations?
model4b <- felm(data = scores16, judge ~ relevel(countrybias,ref = "NON") | jumpid + judge_name)
summary(model4b)

# HOME FIELD BIAS MODELS WITH 2016 DATA
# 
home16 <- season2016 %>% 
    select(competitorName, place, country, competitorNation, lenpts = lengthPoints) %>% 
    mutate(competitionplace = str_sub(place, start=1,end=-6)) %>% 
    mutate(athome = if_else(country==competitorNation,1,0))

# Model 5: Length model
model5b <- felm(data=home16, lenpts ~ athome + competitionplace | competitorName)
summary(model5b)

# Model 6: Style points model
model6b <- felm(data=scores16, judge ~ athome + competitionplace + judge_name + speed + length| competitorName)
summary(model6b)

@@

We contribute to the literature on nationalistic and home field advantage biases in several ways. First, we analyse nationalistic bias and assess to what extent such bias can be explained by unobserved characteristics of the performance. Second, we consider variations in such bias, and discuss this variation in light of sociological theory of evaluation and valuation. Third, by using our data from two time periods and comparing our results to those obtained with similar data but from the markedly different context of the Olympic games, a highest-stakes-contest [@zitzewitz_nationalism_2006], we can shed light on the role of different institutional and incentive contexts and changes over time in the magnitude of bias. Fourth, we study the social and physical ramifications of home advantages.

@@

Our reading of the empirical literature on nationalistic and other evaluative biases suggests there is evidence for such biases in a variety of sports. Given its existence, what are the root causes of such bias and how do they emerge? Theoretically, we can concieve of bias in at least three different ways. We denote these three mechanisms the social psychological mechanism, the cultural legitimacy mechanism and the differential professionalization mechanism, and discuss each of them in turn below.

@@

The second way of theorizing bias is to consider the role of judges and their efforts as evaluation work. Evaluative work and the justification for evaluations as a sociological study has been profoundly influences by the works of Boltanski \& Thevenon [@bolstanski_justification_1991]. A central tenet in their sociology of evaluations is the room for evaluation.

@@

Third, the competition rules also limit the influence of such biases by removing the best and worst of the five scores from the total score. This certainly limits the magnitude of any bias, but as previous research have found significant biases [@zitzewitz_nationalist_2006], it does not seem to root it completely out.

@@

Summing up, it seems meaningful to probe the existence of nationalist biases in different countries, but also study whether nationalist biased have decreased in magnitude over time, due to the general tendency towards increased  professionalization in sports.

@@

## The extant literature on home field advantages

Home field advantage is another type of potential bias in ski jumping. Such advantages, where some competitors benefit from advantages from competing on the home field, may arise in many different ways. Several of the mechanisms that may produce home field advantages are pscyhological sources of superior performances, while others involve judges overrating home field competitors. This point illustrates that the mere presence of a biases is not necessarily illegitimate. In order to improve scoring and fairness in sports, one must understand the mechanisms behind them.

Some research into the home-field advantage has attempted to move our understanding of the causal mechanisms behind  the advantages. Carmichael and Thomas [-@carmichael_homefield_2005] outline the four most common explanations of this phenomena using soccer as an example. The first is a familiarity explanation – the home team is more familiar with the field, the stadium, the locker room, and so forth, which puts the players more at ease in the situation, allowing them to concentrate on performing at their best. Second, traveling to another field could result in fatigue, invoke various stressors, and cause disruptions in pre-competition routines for the away team. All these factors may lead to inferior performances by the away team, relative to their likely performance at the home field. Third, specific rules may favour the home team, giving them the upper hand. Last, there could be a crowd effect. The size of the crowd at the home field might improve individual or team performance. It might also make the judge to be more lenient for the home team, through psychological mechanisms.

Page and Page [-@page_evidence_2010] attributes some of the crowd effect to crowd noise, noting that the judge is under both time pressure (making a correct decision in a short time) and social pressure (from the crowd). A study by Nevill et al. [-@nevill_influence_2002] showed that crowd noise was related to the home-field advantage. Setting up an experiment with one group of judges watching a football match with sound, and another group without, they found that the group watching with sound called 15.5% fewer fouls against the home team than the group who watched in silence. This could also indicate that judges use crowd noise as a guideline in their judging work. The crowd effect may also take another direction, as Anders and Rotthoff [-@anders_homefield_2014] find that football clubs with notoriously violent fans have a larger home-field advantage than peaceful ones. Here, the theory is that the prospect of fans-to-judge violence causes a more lenient judge.

Research on the home-field advantage has also been done in various winter sports. Balmer et al. [-@balmer_home_2001] analysed home-field advantages the winter Olympics from 1908 to 1998, and found some evidence for it in figure skating, freestyle skiing, ski jumping, alpine skiing and short track speed skating. The authors examined the relevance of three of the explanations listed above (familiarity with local conditions, time zones and crowd effect), and found evidence for home condition familiarity.  Conditional on the time zones travelled, travel time as such did not matter. Most evidence, however, was found for the crowd effect theory, as sports that were judged only subjectively had a bigger home-field advantage than sports in which assessments were based on both subjective and objective criteria.

@@

A third hypothesis concerns variation in bias across countries. Nationalistic biases may vary across countries due to cultural-professional variations, where the strength of the bias is lower in countries with strong  jumping cultures and higher in countries where the sport is more peripheral in the larger sports field.

Our fourth hypothesis states that nationalistic biases in ski jumping evaluations decrease over time, possibly as a results of continued professionalization of the sport.

For home field advantages, we propose an additional hypothesis that home field advantages are measurable and impacts athletes performances. Finally, we distinguish between the social aspects of competing on the home field and the physical aspects of competing on the home field, and hypothesize that the physical home field advantage is less salient than the social home field advantage, and thus less important for performances.

@@

To test the hypotheses proposed above, we use data from the FIS World Cup competitions in ski jumping from the 2006-2007 and 2007-2008 seasons as well as from the 2015-2016 seasons. The original data for 2006-2007 and 2007-2008 seasons were downloaded from the FIS web site, coded, edited for errors, and collated into complete data set for analysis. For the season 2015-2016, we wrote automated web scraping software that identified relevant PDF files on the FIS servers, downloaded these files, and scraped them according to our pre-specified instructions. The software used to download and scrape the PDF files are documented in the GitHub repository associated with this article.

@@

## Estimation of nationalistic biases

Following the general approach used in most studies of nationalistic biases (in particular, @emerson_assessing_2009; @zitzewitz_nationalism_2006), we analyse systematic tendencies in scores by individual judges and by the judges’ nationality. We begin with a simple model and add complexity in subsequent steps to approach a stricter test of nationalistic bias. In the first, simple, model, a dummy variable $\phi(I=J)$ indicates whether the judge comes from the same country (J) as the athlete (I) or not. The model can be written as

@@

where $s_{ijp}$ is the score given by judge $j$, on performance $p$, by a certain athlete $i$, and $\epsilon_{ijp}$ is the error term. The $B$ coefficient in this model estimates the difference in expected score if the athlete and the judge come from the same country.

However, such a model does not take into account relevant differences between nations. A Norwegian judge may give higher scores to a fellow countryman’s performance due to a correspondence of their definitions of valuable stylistic elements, in accordance with national variations in evaluation practices.

@@

where $q_p$ is the jump fixed effect. This model will not confuse differences in stylistic culture with nationalistic biases, comparing the potentially biased judge against the four other judges for each jump. Following Zitzewitz (2006), we also estimated a model with so-called compensating biases, relaxing the assumption of objectivity with the four remaining judges. This model allows for the remaining judges to compensate a potentially biased judge, by lowering their scores when they know that one of the others will up his score. We did not find any significant compensating bias in our data, but this particular model did not converge and we thus abstain from reporting it in our set of results.

A limitation of the second model is that it does not capture any inherent leniency the judges may display. If a judge consistently scores performances higher or lower than other judges, then we cannot be sure that the observed bias is not just a particular judge always scoring lower or higher than the average judge. To control for this, we add a judge fixed effect, $l_j$, which leads to our third model:

@@

This model is an improvement over the previous ones, but limitations still remain. The model assumes that bias is constant for each judge across all judges from all nations. Ideally, we would want to allow for variations in bias between individual judges or groups of judges by interacting the judge fixed-effect with the indicator function for judge and competitor nation correspondence. This would in principle let us obtain individual bias estimates for each judge, but our data unfortunately do not allow for such a specific estimation. They do however allow for estimating an interaction between nationalistic bias and the home country of the judge. This leads to the model:

\begin{equation}
s_{ijp} = B∙(L_{j}∙\phi(I=J)) + q_p + \epsilon_{ijp}
\end{equation}

where $L_j$ is an added judge country fixed-effect. In this fourth model, the estimates for B provide an assessment of the average bias in scores by judges from different countries.

## Estimation of home field advantage

To analyse home field advantage we need to separate the problem of bias into the physical and the social domains. That is, whether an athlete actually performs better on the home field or whether he gets better scores by the judges due to social pressure or other factors. Thanks to the natural separation of style scores and length scores in ski jumping, this can be uncovered using similar models.

In order to observe a home field effect in points, be it length points or the judges’ style points, we have to compare scores on a home field with scores on away fields – leading to the assumption that a jumper is somewhat consistent in performance throughout a season. Previously, we used the four other judges as a contrast measure of performance quality. For home field advantage we have to assume that an athlete maintains a relatively constant quality of jumping throughout the season, using his away jumps as a guide to his general performance level. This can be a bold assumption, given that local weather conditions and individual athlete’s “daily form” also may affect these scores, as well as the tiny margins of timing required to “push off” at the right moment. Another problem is that countries with a strong tradition for ski jumping will have both stronger jumpers, host competitions more often due to existing infrastructure, and have more judges in the judge panel.

Again, we construct a dummy-variable using a function, $ω(I=H)$, which will take the value 1 if the jump takes place in country I, the athlete’s home country, and H being the competition country. We then wish to estimate the following model:

\begin{equation}
s_{irs}' = \~B' ∙ω(I=H) + a_{i} +λ_{r} +\epsilon_{ir}
\end{equation}

where $s_{irs}$ are the length points of the athlete, in a specific season, in a specific competition. $\~B'$ is the estimated home field effect in points and $a_i$ is is a fixed effect for the athlete-season combinations. The fixed effect also removes any effect of individual jumpers that “strong jumpers” from traditionally strong ski jumping nations by out estimating each athlete, in each season, separately. Lastly, taking into account the different sizes of the hills, which may result in a skewness that the K-point system fails to adjust for, we add a fixed effect for the hill represented by $λ_r$.

For the social part of the home field advantage we wish to estimate a similar model where style points awarded by the judges serves as the dependent variable.

\begin{equation}
s_{ijprs} = \~B ∙ ω(I=H) + a_{is} + λ_r + l_j + βx_p + \epsilon_{ijprs}
\end{equation}

In this model, we add fixed effects for judges as in (3) to capture leniency of individual judges. We cannot use jump fixed-effects to capture the inherent objective quality of the jump, because home advantage is invariant within a jump, but we can include two central jump characteristics in our data, namely length and speed. This is because “all else being equal, a ski jumper with the same take-off speed but better form while airborne will travel further” [@zitzewitz_nationalism_2006, 78]. We therefore add these two variables to the model, where they and their coefficients are represented by the $βx_p$.

In accordance with norms of open social science, we offer our data and analysis log in an open repository at the time of publication.

@@

@@

Our analysis of nationalistic bias proceeds in four steps. Table 2 presents the results from the first three estimations.

>>>>

---
title: "NATIONALIST BIAS IN SPORT PERFORMANCE EVALUATIONS"
subtitle: "An example from the ski jumping World Cup"
runhead: Judging biases in skijumping
author:
- affiliation: Department of Medical Statistics, University of Oslo
  name: Leiv Tore Salte Rønneberg
- affiliation: Department of Sociology and Human Geography, University of Oslo.
  name: Torkild Hovde Lyngstad
- affiliation: European University Institute
  name: Juho Härkönen
date: "14 september 2017"
keywords: nationalist bias, skijumping, judges, fixed effects
abstract: "High stakes and prizes in international sports competitions necessitate efficiently identifying legitimate winners. Illegitimate biases of non-independent judges represent a major threat to the legitimacy of international competitions. Judges’ nationalities represent one source of bias, as they may prefer athletes of their own nation. Judges may also favour athletes competing within their home countries, creating another potential bias. Such biases, conscious or unconscious, may affect the outcomes of competitions. The literature offers no complete consensus on the magnitude of such biases, hence not whether they affect competition outcomes or not. In this paper, we shed light on these problems, using international ski jumping as an example. We draw on data from the FIS World Cup competitions in the 2006-2008 and 2015-2016 seasons, and employ several fixed-effects based regression designs. Our results reaffirm suspicions of national bias in major ski jumping competitions, but also show their magnitude is too little to be of major relevance to competition outcomes. There is a trend towards smaller nationalist biases over time, which suggests that increasing professionalization of sports also may limit the influence of subjective and nationalist impulses."
thanks: "The authors are grateful for financial support from the Småforsk programme at the University of Oslo, and to Marte Knutsson for research assistance in the early phase of establishing the data set.  At publication time, all code and material used to produce this paper (with the exception of the proprietary data) will be available. The authors are also grateful to Matti Nykänen for being an enduring source of inspiration in many aspects of our lives."
output:
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
    template: /home/torkildl/Dropbox/Setups/thl/rmd-xelatex-template-ms.tex
  html_document: default
bibliography: /home/torkildl/Dropbox/Literature/thl-main-library.bib
anonymous: yes
fontsize: 12pt
geometry: margin=2cm
header-includes:
    - \usepackage{placeins}
    - \usepackage{dcolumn}
    - \usepackage{tabularx}
    - \usepackage{lscape}
    - \usepackage{caption}
fig_caption: false
fig.align: "center"
fig.width: 5
---

@@

### SEASON 2005-06 DATA FROM EARLY ANALYSIS
# Read in data
jumps <- as.data.frame(read_dta(here("data_prepared.dta"))) %>%
    select(order(names(.))) %>%
    arrange(jumpid)
tbl_df(jumps)

@@

### Correct judge name spellings
pairs <- matrix(c("YERLY James","YERRLY James","ASPHAUG Kolbjoern","ASPHAUG Kolbjoer","DALLE AVE Sandro","DALLE Ave Sandro","EGLOFF Ernst","EGLOFF Emst","KARJALAINEN Pirjo","KARJALAINEN Pirvo","KUUSINEN Eero","KUUSINE Eero","LOENG Geir","LOENG Geir S.","LORENZ Michael","LORENTZ Michael","NYMAN Tom","NYMAN Torn","SCHMIEDER Jorg","SCHMIEDER Joerg","SCHMIEDERER Konrad","SCHIMIEDERER Konrad"), ncol=2, byrow=TRUE)
pairs[1,1]
for (x in 1:nrow(pairs)) {
    pair <- pairs[x,]
    j <- j %>% mutate(judge_name = ifelse(judge_name==pair[2],pair[1],judge_name))
}

### Various data preparation
d <- j %>% 
  # Correct erroneous nation coding
  mutate(judge_nation = ifelse(judge_nation=="JAP", "JPN", judge_nation)) %>%
  # Convert to factors
  mutate(judge_name = as.factor(judge_name)) %>%
  mutate(judge_nation = as.factor(judge_nation)) %>%
  mutate(country = as.factor(country)) %>%
  # Generate indicator of national bias
  mutate(natbias = ifelse(competitornation==judge_nation,1,0)) %>%
  # Generate indicator of own nation represented in judge panel 
  group_by(jumpid) %>%
  mutate(represented = max(natbias)) %>%
  ungroup %>%
  # and indicator of compansation bias by other judges
  mutate(compensating = represented - natbias) %>%
  # Generate home field advantage dummy
  mutate(athome = ifelse(competitornation==country,1,0)) %>%
  # Generate indicator of national biar per country
  mutate(countrybias = ifelse(natbias==1,as.character(judge_nation),"NON")) %>%
  mutate(countrybias = as.factor(countrybias)) %>%
  # Home field advantage variables
  mutate(athome = ifelse(competitornation==country,1,0)) %>%
  mutate(competitorname = tolower(competitorname)) %>%
  mutate(athseason = as.factor(str_c(competitorname,as.character(season)))) %>%
  mutate(judge = as.numeric(judge)) %>% 
  # Drop unneeded variables
  select(everything())

#
# Season 2015-2016 analysis
#
season2016 <- data.table::fread(here("scrape/scraped-data-season-2016.csv"), stringsAsFactors = F)
scores16 <- season2016 %>% 
    select(contains("judge"), competitornation = competitorNation, country, place, speed, length, competitorname = competitorName, gender) %>%
    mutate(jumpid = row_number()) %>%
    gather(key=jud, value=val, contains("judge")) %>%
    mutate(info = case_when(str_detect(jud, "Name") ~ "judge_name",
                            str_detect(jud,"Nation") ~ "judge_nation",
                            TRUE ~ "judge")) %>%
    mutate(ltr = str_sub(jud, start=-1)) %>%
    select(-jud) %>%
    group_by(jumpid, ltr) %>%
    arrange(jumpid, ltr) %>% 
    spread(key=info, value=val) %>% 
    ungroup %>% 
    mutate(natbias = if_else(competitornation==judge_nation,1,0)) %>%
    mutate_at(c("judge"), as.numeric) %>% 
    mutate(countrybias = factor(ifelse(natbias==1,as.character(judge_nation),"NON"))) %>% 
    mutate(competitionplace = str_sub(place, start=1,end=-7)) %>% 
    mutate(athome = if_else(country==competitornation,1,0)) %>% 
    mutate(competitorname = tolower(competitorname))

### Combine seasons
###
###
scores06 <- setNames(d, tolower(names(d))) 
scores16 <- setNames(scores16, tolower(names(scores16)))

all_scores <- bind_rows(scores06,scores16, .id="whichseason") %>% 
    select(whichseason, jumpid, judge, judge_name, judge_nation, natbias, countrybias, competitionplace, competitorname) %>% 
    mutate(theseason = if_else(whichseason==1, "2005-06","2015-16")) %>% 
    mutate(seasondummy = if_else(whichseason==1,0,1))

# ESTIMATIONS: NATIONAL BIAS MODELS
#
# Model 1, Naïve model of national bias
model1 <- lm(data = all_scores,judge ~ natbias)
summary(model1)

# Model 2: introducing jump fixed effects
model2 <- felm(data = all_scores, judge ~ natbias | jumpid)
summary(model2)

# Model 2 with judge fixed effects for looking at leniency
model2b <- felm(data = all_scores, judge ~ natbias | jumpid + judge_name)
summary(model2b)

# Model 3: Does national bias vary among nations?
all_scores$countrybias_ord <- relevel(as.factor(all_scores$countrybias), ref="NON")
model3 <- felm(data = all_scores, judge ~ relevel(countrybias_ord,ref = "NON") | jumpid)
summary(model3)

# Model 4a: Estimating changes over time
model4a <- felm(data = all_scores, judge ~ natbias + natbias*seasondummy + seasondummy| jumpid)
summary(model4a)

# Model 4b: Estimating changes in leniency and national variance

## Alternative model 4 with individual-level bias estimates. Not included in the final paper.
model4_indjudges <- felm(data=all_scores, formula=judge ~  natbias*judge_name | jumpid) 
model4_indjudges %>% 
    tidy %>% 
    mutate(natbias = if_else(str_detect(term,"natbias"),"interaction","main")) %>% 
    mutate(judge_name = str_sub(term, start=1+((natbias=="interaction")*8))) %>% 
    mutate(judge_name = if_else(judge_name=="", "overall",judge_name)) %>% 
    select(natbias, judge_name, estimate) %>% 
    group_by(judge_name) %>% 
    spread(key=natbias, estimate) %>% 
    rename(bias = interaction, leniency = main) %>% 
    ungroup %>% 
    mutate(judge_name = str_sub(judge_name,start=11)) %>% 
    head

@@

We contribute to the literature on nationalistic biases in several ways. First, we assess to what extent nationalistic bias can be explained by unobserved characteristics of the performance. Second, we consider national variations in such bias, and discuss this variation in light of sociological theory of evaluation and valuation. Third, by using our data from two time periods and comparing our results to those obtained with similar data but from the markedly different context of the Olympic games, a highest-stakes-contest [@zitzewitz_nationalism_2006], we can shed light on the role of different institutional and incentive contexts and changes over time in the magnitude of bias.

@@

Our reading of the empirical literature on nationalistic and other evaluative biases concludes that there is evidence for such biases in a variety of sports. Given its existence, what are the root causes of such bias and how do they emerge? Theoretically, we can concieve of bias resulting from at least three different mechanisms. We denote these three mechanisms the social psychological mechanism, the cultural legitimacy mechanism and the differential professionalization mechanism, and discuss each of them in turn below.

@@

The second way of theorizing bias is to consider the role of judges and their efforts as evaluation work. Evaluative work and the justification for evaluations as a sociological study has been profoundly influences by the works of Boltanski \& Thevenon [@boltanski_justification_1991]. A central tenet in their sociology of evaluations is the room for evaluation.

@@

Third, the competition rules also limit the influence of such biases by removing the best and worst of the five scores from the total score. This certainly limits the magnitude of any bias, but as previous research have found significant biases [@zitzewitz_nationalism_2006], it does not seem to root it completely out.

@@

Summing up, it seems meaningful to reexamine the existence of nationalist biases, study how these biases vary across the countries from which evaluators are drawn, and study whether nationalist biased have decreased in magnitude over time, due to, for example, the general tendency towards increased professionalization in sports.

@@

@@

A third hypothesis concerns variation in bias across countries. Nationalistic biases may vary across countries due to cultural, historical and professional ideosyncracies at the national level. The strength of the bias is assumed to be lower in countries without strong ski jumping cultures and higher in countries where the sport is more peripheral in the larger sports field.

Our fourth hypothesis states that nationalistic bias in ski jumping evaluations decrease over time, possibly as a result of the continued professionalization of the sport.

@@

To test the hypotheses proposed above, we use data from the FIS World Cup competitions in ski jumping from the 2006-2007 and 2007-2008 seasons as well as from the 2015-2016 seasons. The original data for 2006-2007 and 2007-2008 seasons were downloaded from the FIS web site, coded, edited for errors, and collated into complete data set for analysis. For the season 2015-2016, we wrote automated web scraping software that identified relevant PDF files on the FIS servers, downloaded these files, and scraped them for results according to our pre-specified instructions. In accordance with norms of open social science, the software used to download and scrape the PDF files are documented in the GitHub repository associated with this article.[^1: See https://github.com/torkildl/skijumping for data and software.]

@@

## Estimation of nationalistic bias using fixed effects models

Following the general approach used in most studies of nationalistic biases (in particular, @emerson_assessing_2009; @zitzewitz_nationalism_2006), we analyse systematic tendencies in scores by individual judges and by the judges’ nationality. We begin with a simple model and add complexity in subsequent steps to test our different hypotheses.

In the first, simple, model, a dummy variable $\phi(I=J)$ indicates whether the judge comes from the same country *j* as the athlete *i* or not. The model can be written as

@@

where $s_{ijp}$ is the score given by judge $j$, on performance $p$, by a certain athlete $i$, and $\epsilon_{ijp}$ is the error term. The $B$ coefficient in this model estimates the difference in expected score if the athlete and the judge come from the same country. Hypothesis \#1 is tested by assessing the direction and magnitude of $B$.

Such a model does not take into account relevant differences between nations. A Norwegian judge may give higher scores to a fellow countryman’s performance due to a correspondence of their definitions of valuable stylistic elements, in accordance with national variations in evaluation practices.

@@

where $q_p$ is the jump fixed effect. This model will not confuse differences in stylistic culture with nationalistic biases, comparing the potentially biased judge against the four other judges for each jump. The $B$ estimate recovered from this model will be compared with the corresponding parameter from model 1, and the result of this comparison can be applied to test hypothesis \#2.

A limitation of the second model is that it does not capture any inherent leniency the judges may display. If a judge consistently scores performances higher or lower than other judges, then we cannot be sure that the observed bias is not just a particular judge always scoring lower or higher than the average judge. To control for this, we add a judge fixed effect, $l_j$, which leads to an adjusted model \#2b:

@@

This model is an improvement over the previous ones, but limitations still remain. For example, there could be so-called compensating biases, where other judges on thethe same panel as a suspected biased judge compensate by lowering their scores when they know that one of the others will up his score. Finally, the model also assumes that bias is constant across all judges from all nations, a crucial assumption that we subsequently question in hypothesis \#3.

## Variations across nations and time periods

In order to test our third hypothesis, we want to allow for variations in bias between groups of judges from the same nation by interacting the judge fixed-effect in model \#2 with the indicator function for judge and competitor nation correspondence. This implies the followung model:

\begin{equation}
s_{ijp} = B_{j}∙(L_{j}∙\phi(I=J)) + q_p + \epsilon_{ijp}
\end{equation}

where $L_j$ is an added judge country fixed-effect. The nation-specific estimates for B provide an assessment of the average bias in scores by judges from different countries.\footnote{In principle, we could obtain individual bias estimates for each judge and then analyze these as samples of judge biases scores grouped by judges' nationalities. This would be very demanding of our data and experiments indicate that estimates would be very imprecise. We thus refrain from doing this, and only allow the bias to vary between the different judge nationalities.} We also rank countries by our selected instrument of ski jumping history, namely the historical record of Olympic ski jumping gold medals in the decades before the seasons we study. The hypothesis is tested by estimating the Spearman rank correlation $\rho$ across the countries. A negative correlation indicates support for the hypothesis.

Our fourth hypothesis regards potential changes in the variation of nationalist bias over time. In order to assess changes in the variation in nationalist bias, we estimate a model akin to model \#2, but with an added interaction with a dummy representing the second season (with the first season as a reference group). The parameter estimate for this interaction effect, $\Gamma$ provides a direct test of hypothesis with regard to changes in nationalist bias. The model is written as

\begin{equation}
s_{ijp} = B∙\phi(I=J) + \Gamma∙Y∙\phi(I=J)+ q_p + \epsilon_{ijp}
\end{equation}

@@

Our analysis of nationalistic bias proceeds in four steps, with each step corresponding to a test of one of our four hypotheses.

@@

Table 2 presents the results from the first three estimations. Our first model yields a positive and statistically significant estimate: judges award an average of 0.146 extra points to jumpers from their home country. This is approximately one-third of the minimum increment in judge scores in ski jumping (0.5). Compared with within-jump variance it only amounts to a quarter of a standard deviation. This result already suggests that the magnitude of nationalistic bias in judges’ scores is limited. The estimate may be biased by, for example, cultural and historical differences between countries that lead to an over- or underappreciation of certain styles.

\begin{landscape}
```{r tab2, echo=FALSE, results='asis'}
stargazer(model1, model2, model2b, out.header=FALSE, header=FALSE,
    single.row=TRUE,
    column.labels = c("Model 1: Naïve", "Model 2a: Within-jump", "Model 3a: Within-judge"),
    title="Results from regression models of national bias",
    covariate.labels = c("National bias"),
    dep.var.caption  = "", dep.var.labels.include = FALSE,
    model.numbers = FALSE, model.names = FALSE,
    add.lines = c(" "),
    notes.align="l",
    notes = c("In models 2 and 2b, the intercept is suppressed by the fixed effects estimation procedure."),
    notes.append=TRUE,
    df = TRUE, decimal.mark =".",
    ci=TRUE, ci.level=0.95, ci.separator = "  ",
    omit.stat=NULL,
    star.cutoffs = c(0.05,0.01,0.001)
    )
```
\end{landscape}
